Linting Linearity in Core/System FC

### 2022-11-26 proposal.tex
Tweaks and small extensions

This document is a work in progress proposal that I will deliver (in
an extended form) at the end of
the semester as part of my master thesis preparation. The document needs a
proper introduction and lengthy background and related work section besides the
content (to be extended) that follows, such that an outsider might understand
what I propose to do. However, this initial iteration is instead targeted at
those already familiar with the problem and serves as a (less bureaucratic)
proposal on linting linearity in Core and to showcase the preliminary progress I
have made so far.

@@

Since the publication of Linear Haskell~\cite{linearhaskell} and its release in GHC 9.0,
Haskell's type system supports linearity annotations in functions -- bringing
linear types to a mainstream, pure, and lazy functional language. Concretely,
function types can be annotated with a multiplicity (a multiplicity of One
requires the argument to be consumed exactly once, a multiplicity of Many allows
the argument to be consumed unrestrictedly, i.e., zero or more times). A
function is linear if it consumes its arguments exactly once when it itself is
consumed exactly once (its arguments have multiplicity of one), for some
definition of \emph{consume} that I should revise here.

@@

Core is a much smaller language than the whole of Haskell (even though we
can compile the whole of Haskell to it!) and its type checker is
simple and fast due to Core being explicitly typed. In essence, Core
is close to a higher-order polymorphic lambda calculus. This type
checker (called \emph{Lint}) gives us guarantees of correctness in face of all
the complex transformations a Haskell program undergoes, such as desugaring and
core-to-core optimization passes, because the linter is always run on the resulting code
before being ultimately compiled to (untyped) machine code.

@@

In this sense, Core is already annotated with linearity, but the \textbf{linter currently
  ignores linearity annotations}.
%
In spite of the strong formal foundations of linear types driving the
implementation, their interaction with the whole of GHC is still far
from trivial, and the implemented type system cannot accomodate
several optimising transformations that produce programs that violate
linearity in a syntactically (i.e.,~multiple occurrences of linear
variables in the program text), but ultimately preserve it in a
semantic sense, where a linear term is still consumed exactly once --
this is compounded by lazy evaluation driving a non-trivial mismatch
between syntactic and semantic linearity. Therefore, linear linting
rejects various valid programs (with regard to linearity) after
desugaring. The current solution to this is to effectively disable the
linear linter, since disabling optimisations can incur significant
performance costs. However, we believe that GHC's transformations are
correct, and it is the linear linter and its underlying type system
that that cannot sufficiently accommodate the
resulting programs.
% diverse problems with linearity spring up when we get past the desugarer and
% the
% linter rejects valid programs which is quite undesirable.

@@

We propose to extend Core's type system with additional linearity
information in order to to accommodate linear program in Core that
result from optimising transformations. We will prove the system's
soundness/type safety and show how it validates some core-to-core optimisation
passes. We also plan to implement the extension into GHC's
linter. Concretely, we will extend Core's linear type system with
\emph{usage annotations} for let, letrec and case binder bound
variables. Additionally, we will explore a new kind of coercion for
multiplicities to validate programs that combine GADTs with
multiplicities.
%
The ultimate measure of success is the \verb=-dlinear-core-lint= flag,
which activates the linear linter. In its current implementation,
enabling this flag rejects many linearly valid programs. Ideally, by
the end of our research and implementation, this flag could be enabled by
default and accommodate all existing transformations. Realistically, we want to
accept as many diverse transformations as possible while still preserving
linearity, even if we are unable to account for all of them.

@@

Despite not being accepted by the surface-level language, linear programs using
lets occur in Core due to optimising transformations that create let
bindings. In a similar fashion, programs which violate syntatic linearity
for other reasons other than let bindings are produced by Core transformations.

@@

Currently, every variable in Core is annotated with a multiplicity at its binding
site. The multiplicity for let-bound variables must be ignored throughout the
transformation pipeline or otherwise too many programs valid programs would be
rejected for violating linearity in its transformed type.

@@

We describe below an algorithm for computing the usage environment of a set of
recursive definitions, and its textual description. An (at least informal) proof
that the algorithm is correct should eventually follow this.

@@

The current typing rule for case expressions describes that using the case
binder is as using the case scrutinee multiple times (though it seems like it
actually just starts counting after the second usage which would make the rule
wrong for a case expression that doesn't use the binder -- the scrutinee should
also be consumed), so we scale the usage of
the case scrutinee by the superior multiplicity of using the case binder across
all branches. However, we hypothetize that this prevents us from typing many
valid programs and ultimately does not correctly capture the usage of a variable,
and present a seemingly viable alternative.

@@

Now, with the following example, my interpretation of the current rule says that
the following case expression consumes the scrutinee twice because the case
binder is used. However, we know that in the branch where z is used, z
is equal to False, and using False does not violate linearity since we can
unrestrictedly create nullary data constructors. A situation similar to this one
below happens after the CSE transformation.
\begin{code}
    case <complex expression> of z {
        False -> case y of z' { DEFAULT -> z }
        True  -> y
    }
\end{code}
% $\leadsto U$

@@

The key idea of the case-binder-usage solution is annotating the case binder
with independent usage environments for each pattern match. That is, for
each of the possible branches, using the case binder $z$ will equate to
using its usage environment in that branch. That makes it so that using the
case binder instead of a nullary data constructor is the same, using the case
binder instead of reconstructing the value with the bound pattern variables too,
and other situations in which we previously couldn't typecheck linearity are
now accepted.

@@

>>>>

This document is a work in progress proposal that I will deliver (in an extended
form) at the end of the semester as part of my master thesis preparation. The
document still needs a proper introduction, lengthy background and related work
section  such that an outsider might understand what I propose to do. However,
this initial iteration is instead targeted at those already familiar with the
problem and serves as a (less bureaucratic) proposal on linting linearity in
Core and to showcase the preliminary progress I have made so far.

@@

Since the publication of Linear Haskell~\cite{linearhaskell} and its release in GHC 9.0,
Haskell's type system supports linearity annotations in functions -- bringing
linear types to a mainstream, pure, and lazy functional language. Concretely,
function types can be annotated with a multiplicity (a multiplicity of One
requires the argument to be consumed exactly once, a multiplicity of Many allows
the argument to be consumed unrestrictedly, i.e., zero or more times). A
function is linear if it consumes its arguments exactly once when it itself is
consumed exactly once.
% TODO: for some definition of \emph{consume} that I should revise here.

@@

Core is a much smaller language than the whole of Haskell (even though we
can compile the whole of Haskell to it!) and its type checker is
simple and fast due to Core being explicitly typed. In essence, Core
is close to a higher-order polymorphic lambda calculus. This type
checker (called \emph{Lint}) gives us guarantees of correctness in face of all
the complex transformations a Haskell program undergoes, such as desugaring and
core-to-core optimization passes, because the linter is always run on the resulting code
before ultimately being compiled to (untyped) machine code.

@@

In this sense, Core is already annotated with linearity, but the \textbf{linter currently
  ignores linearity annotations}.
%
In spite of the strong formal foundations of linear types driving the
implementation, their interaction with the whole of GHC is still far
from trivial. The implemented type system cannot accomodate
several optimising transformations that produce programs that violate
linearity syntactically (i.e.,~multiple occurrences of linear
variables in the program text), but ultimately preserve it in a
semantic sense, where a linear term is still consumed exactly once --
this is compounded by lazy evaluation driving a non-trivial mismatch
between syntactic and semantic linearity. Therefore, linear linting
rejects various valid programs (with regard to linearity) after
desugaring. The current solution to this is to effectively disable the
linear linter, since otherwise disabling optimisations can incur significant
performance costs. However, we believe that GHC's transformations are
correct, and it is the linear linter and its underlying type system
that cannot sufficiently accommodate the
resulting programs.
% diverse problems with linearity spring up when we get past the desugarer and
% the
% linter rejects valid programs which is quite undesirable.

@@

We propose to extend Core's type system with additional linearity
information in order to accommodate linear programs in Core that
result from optimising transformations. We will prove the system's
soundness/type safety and show how it validates some core-to-core optimisation
passes. We also plan to implement the extension into GHC's
linter. Concretely, we will extend Core's linear type system with
\emph{usage annotations} for let, letrec and case binder bound
variables. Additionally, we will explore a new kind of coercion for
multiplicities to validate programs that combine GADTs with
multiplicities.
%
The ultimate measure of success is the \verb=-dlinear-core-lint= flag,
which activates the linear linter. In its current implementation,
enabling this flag rejects many linearly valid programs. Ideally, by
the end of our research and implementation, this flag could be enabled by
default and accommodate all existing transformations. Realistically, we want to
accept as many diverse transformations as possible while still preserving
linearity, even if we are unable to account for all of them.

@@

Despite not being accepted by the surface-level language, linear programs using
lets occur naturally in Core due to optimising transformations that create let
bindings. In a similar fashion, programs which violate syntatic linearity
for other reasons other than let bindings are produced by Core transformations.

@@

Currently, every variable in Core is annotated with a multiplicity at its binding
site. The multiplicity for let-bound variables must be ignored throughout the
transformation pipeline or otherwise too many valid programs would be
rejected for violating linearity in its transformed type.

@@

The algorithm for computing the usage environment of a set of
recursive definitions works as follows. An at least informal proof
that the algorithm is correct should eventually follow this.

@@

% (though in the rule seems to only start counting after the second usage which
% would make the rule wrong for a case expression that doesn't use the binder --
% the scrutinee should also be consumed)
The current typing rule for case expressions describes that using the case
binder is as using the case scrutinee multiple times, so we scale the usage of
the case scrutinee by the superior multiplicity of using the case binder across
all branches. However, we hypothetize that this prevents us from typing many
valid programs and ultimately does not correctly capture the usage of a
variable and present a seemingly viable alternative.

@@

Now, with the following example, my interpretation of the current rule dictates
that the case expression consumes the scrutinee twice because the case binder is
used. However, we know that in the branch where $z$ is used, $z$ is equal to
False, and using False does not violate linearity since we can unrestrictedly
create nullary data constructors. A situation similar to this one below happens
after the CSE transformation.
\begin{code}
    case <complex expression> of z {
        False -> case y of z' { DEFAULT -> z }
        True  -> y
    }
\end{code}
% $\leadsto U$

@@

The key idea of the case-binder-usage solution is \textbf{annotating the case binder
with independent usage environments for each pattern match}. That is, for
each of the possible branches, using the case binder $z$ will equate to
using its usage environment in that branch. That makes it so that using the
case binder instead of a nullary data constructor is the same, using the case
binder instead of reconstructing the value with the bound pattern variables too,
and other situations in which we previously couldn't typecheck linearity are
now accepted.

@@

Taking the first example in this section, we annotate both branches (True and
False) with $U_z = \emptyset$ because the pattern match doesn't introduce any
new variables. We consume \emph{complex expression} once and $y$ once, and then use
$z$ once. However, using $z$ once equates to not using anything (semantically
this relates to being able to use False or True unrestrictedly) and therefore
linearity is preserved.

In the second example, we know annotate the only branch with usage environment
$U_z = [x := 1, y := 1]$. Upon the usage of $z$ we emit $x$ and $y$ --
effectively using both -- thus linearity is preserved.
